---
title: "yeg_opioid_report"
author: "Mike Zhang"
format: 
  html:
    toc: true
    toc-depth: 3
    nunmber-sections: true
    theme: cosmo
    embed-resources: true
    output-file: index.html
    output-dir: "/docs"
execute:
  echo: false
  warning: false
  message: fakse
---

```{r}
#| label: set-up

library(tidyverse)
library(lubridate)
library(gtsummary)
library(nlme)

edmonton_opioid <- readRDS("data/processed/edmonton_opioid.rds")
calgary_opioid<- readRDS("data/processed/calgary_opioid.rds")
edmonton_any <- readRDS("data/processed/edmonton_any.rds") 
calgary_any<- readRDS("data/processed/calgary_any.rds")

edmonton_opioid_final <- readRDS("models/edmonton_opioid_final.rds")
calgary_opioid_final <- readRDS("models/calgary_opioid_final.rds")
edmonton_any_final <- readRDS("models/edmonton_any_final.rds")
calgary_any_final <- readRDS("models/calgary_any_final.rds")

```

# Introduction

The implementation of supervised consumption sites (SCS) has been a contentious and heavily politicized public health topic in Alberta over the past few years. With the Alberta government having already closed some SCSs within the province and planning further closures in the near future, this project examines the association between SCS implementation and drug-related death rates in two major Alberta cities, Edmonton and Calgary, using segmented linear regression on time series data of opioid and all-substance death rates per 100,000 population. All data were scraped from the Alberta Substance Use Surveillance System and span January 2016 to March 2026. The main findings complicate a simple narrative because no statistically significant effect of SCS implementation was found in Calgary, for either opioid or all-substance death rates. In Edmonton, however, both opioid and all-substance death rates rose significantly faster following SCS implementation relative to control zones, and the near-identical effect across both outcomes suggests this reflects factors specific to Edmonton during this period rather than a mechanism attributable to SCS itself.

# Background

Supervised consumption sites (SCS) are health facilities where individuals can use pre-obtained drugs under the observation of trained staff, who can intervene with oxygen or naloxone if an overdose occurs. Naloxone specifically reverses opioid overdoses by blocking opioid receptors. It has no equivalent mechanism for cocaine, methamphetamine, alcohol, or most other substances. This mechanism matters for interpreting this project's findings, discussed further below.

Alberta's first SCS opened at Calgary's Sheldon M. Chumir Health Centre in October 2017, followed shortly after by sites in Edmonton (Boyle Street Community Services in March 2018, George Spady Centre in April 2018, and Boyle McCauley Health Centre in November 2018) and the Royal Alexandra Hospital's in-patient program. Over the following two years, sites also opened in several other Alberta cities: Lethbridge's ARCHES site opened February 28, 2018 (closed by the provincial government in 2020, later replaced by a separate site at the Lethbridge Shelter), Red Deer opened a site around September 2018 (closed March 2025), and Grande Prairie's mobile site opened March 11, 2019 and remains in operation. Lethbridge, Red Deer, and Grande Prairie fall within Alberta's South, Central, and North health zones respectively — the same three zones used as comparison groups in this analysis, a point revisited in the Methods section.

SCS implementation has remained politically contentious in Alberta since these initial openings, with debate centered around concerns about crime, social disorder, and needle debris in surrounding neighbourhoods, weighed against harm-reduction outcomes such as overdose reversals and reduced disease transmission. Under the current provincial government, policy has shifted decisively toward a "recovery-oriented" model emphasizing treatment and detox services over harm reduction. This shift has translated into concrete closures: the Royal Alexandra Hospital site in Edmonton closed in December 2025, and in March 2026 the government announced the closure of the Calgary Chumir site and the Lethbridge site, both effective June 30, 2026 — leaving only two sites in Edmonton and the Grande Prairie mobile site remaining in the province. The government has cited a report from the Canadian Centre of Recovery Excellence, a Crown corporation it funds, showing no increase in mortality following the Red Deer site's 2025 closure; critics have challenged both the report's independence and its conclusions, and some have pointed to rising drug poisoning deaths in Lethbridge following the 2020 closure of the ARCHES site as evidence of harm.

The broader evidence base on SCS effectiveness is itself mixed. International research, most notably on Vancouver's Insite facility, has generally reported reductions in overdose mortality and safer injection practices in surrounding areas. Evidence specific to Alberta and other Canadian jurisdictions is comparatively limited and contested, with government-commissioned reviews and independent research sometimes reaching different conclusions. Given the scale of the closures currently underway and the province's stated intention to move away from this model entirely, a locally-scoped, methodologically rigorous evaluation of SCS impact in Alberta's two largest cities is timely.

# Method

## Data

All the data analyzed were scraped from Alberta Substance Use Surveillance System (ASUSS)'s SAS VA dashboard under its "Acute substance deaths overview" tab using Google Chrome Developer Tools' Network tab to locate and download the underlying CSV export files. The monthly rates of drug poisoning deaths per 100,000 person years of all five AHS zones ranged from January 2016 to March 2026. Death rates of both any opioid and any substance were extracted as outcome variables for specificity check later. Among the zones, Edmonton and Calgary zones were modeled as treatment zones, while the other three (Central, North, South) were treated as control zones. For intervention starting dates, March 1st 2018 and October 1st 2017 were chosen for Edmonton and Calgary zones, respectively, because these were the months where each zone's first SCS opened, and no denominator information was provided in its accompanying documentation.

## Study Design

Controlled interrupted time series (ITS) was selected for this project as it fits our scenario where a known-date intervention with a comparison group is available. Since SCS intervention dates were different for Edmonton and Calgary zones, they were modeled separately instead of pooled, which would force one shared date across both treated zones and misrepresent Calgary's and Edmonton's actual staggered timing. Notably, other three zones that served as the control group also opened and closed their own SCS sites over the years, so the analysis is more like a staggered-rollout comparison rather than treated-vs-never-treated.

## Model Specification

The final segmented regression formula is `rate ~ time + zone + intervention + time_after + covid + covid_time` where `rate` represents drug-related death rates per 100,000 person-years, `time` represents months passed since the earliest date of the data, `zone` represents AHS zones, a binary `intervention` variable indicates whether SCS was in effect or not, `time_after` serves as months passed since the intervention dates (0 before the intervention, and for control zones throughout), another binary `covid` variable indicates whether COVID-19's lockdown, arbitrarily set to March 1st 2020, has happened, and `covid_time` represents months elapsed since COVID-19.

COVID-19 was modeled with both a level shift (`covid`) and a slope change (`covid_time`) rather than a level shift alone. This decision followed a specific diagnostic sequence where a formal linearity check (adding a quadratic term to `time_after` and testing it via likelihood ratio test) found significant curvature in the post-intervention segment for both Edmonton and Calgary. Visual inspection suggested this curvature coincided with the onset of COVID-19 rather than reflecting a genuine nonlinear SCS effect. To test this, `covid_time` was added to allow COVID-19 its own slope, separate from `time_after`. This resolved the issue since once `covid_time` was included, the quadratic term was no longer statistically significant, confirming that the apparent curvature was attributable to COVID-19's distinct trajectory rather than non-linearity in the SCS effect itself.

## Auto-correlation

To be able to run segmented linear regression models, autocorrelation among data points at different dates needed to be corrected. An Ordinary Least Squares model without correction would distort the standard errors of parameter coefficients, leading to unreliable significance tests, even though the coefficient estimates themselves would remain unbiased. Generalized Least Squares (GLS) models were utilized over other alternatives, such as Prais-Winsten and Newey-West/HAC, because the data had a panel structure with multiple zones where zone-specific correction was needed. GLS was also favored over Generalized Estimating Equations (GEE), a comparable panel-data alternative, because GLS's likelihood-based framework allowed the autocorrelation correction to be formally tested against a plain OLS model rather than simply assumed. The argument `method = "ML"` was included in `gls()` to fit AR(1)-corrected GLS models on a scale comparable to OLS models, and the two were compared using `anova()`, which returned statistically significant results for both treatment zones, justifying the AR(1) correction. All final models were run with `method = "REML"` for reporting.

## Assumption Checks

### Parallel Pre-trends

GLS models incorporating AR(1) correction using the formula `rate ~ time * zone` were run on each zone's pre-intervention period, bounded by that treated zone's own intervention date (`2017-10-01` for Calgary zone and `2018-03-01` for Edmonton zone). The results showed non-significant interaction terms for both Calgary and Edmonton against all three comparison zones, supporting the parallel pre-trends assumption.

### Linearity

Linearity was assessed both within the post-intervention segment specifically (via a quadratic term on time_after), and across the overall shared time trend (via a quadratic term on time), each tested against the baseline AR(1) model via likelihood ratio test. The post-intervention segment showed no significant curvature for either zone (Calgary: p = 0.256; Edmonton: p = 0.159), supporting a linear specification for the SCS effect itself. The overall trend, however, showed significant curvature for both zones (Calgary: p = 0.0003; Edmonton: p = 0.0002). This was investigated further and attributed to COVID-19's distinct trajectory rather than non-linearity in the SCS effect; see Model Specification for the resolution.

### Residual Diagnosis

QQ plots showed right-skewed residuals in the primary models, likely driven by small monthly death counts in low-population zones. Log-transforming the outcome variable was attempted to address this, but it introduced skewness in the opposite direction, traced to zone-months with a death rate of zero (e.g., South and North zones in several months), and additionally broke the parallel pre-trends assumption for the South zone specifically. Given these issues, the raw-scale model was retained as the primary specification, with the log-transformed model kept only as a secondary robustness check (see Sensitivity and Specificity Analyses). The residual skew in the primary model remains a disclosed limitation, likely reflecting the inherent instability of rate estimates built on small monthly death counts in lower-population zones.

## Sensitivity and Specificity Analyses

### Specificity Check

A specificity check was conducted by swapping the original outcome variable (any opioid death rates) for any-substance death rates. This tested mechanistic specificity to opioids, given SCS's naloxone-based mechanism, which has no plausible pathway to affect non-opioid deaths. Results were nearly identical to the baseline AR(1) models, particularly for Edmonton, suggesting a zone-wide confound rather than an SCS-specific effect.

### Sensitivity Analysis

Two sensitivity analyses were conducted. The first excluded the South zone from the comparison group, given that Lethbridge (South zone) opened its own SCS within weeks of Edmonton's, making South a compromised control. The second fit uncontrolled single-zone models with no comparison zones at all. Both single-zone models showed no statistically significant SCS effect. Under South-exclusion, Calgary's `time_after` coefficient flipped sign and became statistically significant; however, given the number of specifications tested across this analysis, this result should be interpreted with caution due to the risk of inflated Type I error from multiple comparisons. Both sensitivity analyses were conducted after the primary model was finalized, to avoid the appearance of post-hoc specification searching.

## Software

All analyses and visualizations were performed on R version 4.6.1. Packages utilized include `tidyverse` for general data manipulation and visualization (`ggplot2`), `lubridate` for date variables, and `nlme` for GLS modeling.

# Results

## Primary Model Results

```{r}
#| label: tbl-1
#| tbl-cap: "Segmented linear regression estimates for the effect of supervised consumption site (SCS) implementation on opioid-related death rates in Edmonton, with North, Central, and South zones as comparison groups (Central zone serves as the reference category)."

library(broom.mixed)

edmonton_opioid_final |> 
  tbl_regression(
    label = list(
      time         ~  "Overall time passed",
      zoneEdmonton ~  "Edmonton zone",
      zoneNorth    ~  "North zone",
      zoneSouth    ~  "South zone",
      intervention ~  "Immediate change at SCS opening",
      time_after   ~  "Time passed since intervention",
      covid        ~  "COVID-19",
      covid_time   ~  "Time passed since COVID-19"
    )
  ) |> 
  modify_header(label ~ "**Variable**") |> 
  as_gt()

```

``` {r}
#| label: tbl-2
#| tbl-cap: "Segmented linear regression estimates for the effect of supervised consumption site (SCS) implementation on opioid-related death rates in Calgary, with North, Central, and South zones as comparison groups (Central zone serves as the reference category)."

calgary_opioid_final |> 
  tbl_regression(
    label = list(
      time         ~  "Overall time passed",
      zoneCalgary  ~  "Calgary zone",
      zoneNorth    ~  "North zone",
      zoneSouth    ~  "South zone",
      intervention ~  "Immediate change at SCS opening",
      time_after   ~  "Time passed since intervention",
      covid        ~  "COVID-19",
      covid_time   ~  "Time passed since COVID-19"
    )
  ) |> 
  modify_header(label ~ "**Variable**") 

summary(calgary_opioid_final)
```

According to the primary model results, Calgary showed no statistically significant SCS effect on opioid death rates — neither the immediate level shift at SCS opening nor the change in trend afterward reached significance (p = 0.6 and p = 0.3, respectively). Edmonton, by contrast, showed a statistically significant increase in the trend following SCS implementation (p = 0.004), indicating opioid death rates rose faster than expected relative to comparison zones after the SCS opened, though the immediate level shift itself was not significant.

COVID-19, modeled separately from the SCS effect, showed a significant positive immediate level shift in both zones (p < 0.001 for both), alongside a significant negative change in trend afterward (Edmonton: p = 0.019; Calgary: p = 0.026) — consistent with an initial spike in death rates at pandemic onset followed by a gradual moderation in the growth rate over time.

## Trajectory Visualizations

```{r}
#| label: prepare visualization data

edmonton_plot_data <- edmonton_opioid |> 
  ungroup() |> 
  mutate(fitted = fitted(edmonton_opioid_final))
calgary_plot_data <- calgary_opioid |> 
  ungroup() |> 
  mutate(fitted = fitted(calgary_opioid_final))

edmonton_counter_factual <- edmonton_opioid |> 
  mutate(
    intervention = 0,
    time_after = 0,
    time_after_sq = 0
  ) 
calgary_counter_factual <- calgary_opioid |> 
  mutate(
    intervention = 0,
    time_after = 0,
    time_after_sq = 0
  )

edmonton_counter_factual$counter_factual_fit <- predict(edmonton_opioid_final, newdata = edmonton_counter_factual)
calgary_counter_factual$counter_factual_fit <- predict(calgary_opioid_final, newdata = calgary_counter_factual)

edmonton_plot_data <- edmonton_plot_data |> 
  filter(zone == "Edmonton")
calgary_plot_data <- calgary_plot_data |> 
  filter(zone == 'Calgary')
edmonton_counter_factual <- edmonton_counter_factual |> 
  filter(zone == "Edmonton") 
calgary_counter_factual <- calgary_counter_factual |> 
  filter(zone == "Calgary")

```

```{r}
#| label: fig-1
#| fig-cap: "Observed and model-fitted opioid death rates for Edmonton, with a counterfactual trajectory assuming no SCS effect. Gray points represent observed monthly death rates. The solid blue line shows the model's fitted trajectory; the gray dashed line shows the counterfactual trajectory, generated by holding the SCS intervention terms at zero while retaining all other model terms (including COVID-19). The solid amber vertical line marks SCS implementation (March 2018); the dotted line marks COVID-19 onset (March 2020). The widening gap between the fitted and counterfactual lines after 2020 reflects the statistically significant increase in trend following SCS implementation (p = 0.004); the small level shift visible immediately at the amber line is not statistically significant and should not be interpreted as a real effect."

ggplot() +
  geom_point(data = edmonton_plot_data, aes(x = date, y = rate)) +
  geom_line(data = edmonton_plot_data, aes(x = date, y = fitted), color = "#2a78d6", linewidth = 1) +
  geom_line(data = edmonton_counter_factual, aes(x = date, y = counter_factual_fit), color = "#888780", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = ymd("2018-03-01"), linetype = "solid", color = "#eda100", linewidth = 0.8) +
  geom_vline(xintercept = ymd("2020-03-01"), linetype = "dotted", color = "gray50") +
  labs(
    title = "Edmonton: observed opioid death rate vs. model-fitted and counterfactual trajectories",
    x = "Date", y = "Rate per 100,000 person-years"
  ) +
  theme_minimal()

```

```{r}
ggplot() +
  geom_point(data = calgary_plot_data, aes(x = date, y = rate)) +
  geom_line(data = calgary_plot_data, aes(x = date, y = fitted), color = "#2a78d6", linewidth = 1) +
  geom_line(data = calgary_counter_factual, aes(x = date, y = counter_factual_fit), color = "#888780", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = ymd("2017-10-01"), linetype = "solid", color = "#eda100", linewidth = 0.8) +
  geom_vline(xintercept = ymd("2020-03-01"), linetype = "dotted", color = "gray50") +
  labs(
    title = "Calgary: observed opioid death rate vs. model-fitted and counterfactual trajectories",
    x = "Date", y = "Rate per 100,000 person-years"
  ) +
  theme_minimal()
```

The counter-factual trajectories in @fig-1 and @fig-2 were generated by taking each zone's fitted model and setting the SCS-related terms (immediate level shift and trend change) to zero, while holding all other model terms, including the shared time trend, zone baseline, and COVID-19 effect, unchanged. The resulting gap between the fitted and counter-factual lines therefore isolates the modeled SCS effect specifically after accounting for the pandemic-era trajectory shared across all zones. This is visible directly in Edmonton (@fig-1), where the fitted and counter-factual lines diverge steadily after 2020, consistent with the statistically significant increase in post-intervention trend reported above. In Calgary (@fig-2), the two lines remain close throughout the post-intervention period, consistent with the null result found in the primary model.

## Specificity Check

```{r}
#| label: tbl-3
#| tbl-cap: "Comparison of the estimated trend change (`time_after`) across opioid and any-substance outcomes, by zone."

tbl_calgary_opioid  <- tbl_regression(
  calgary_opioid_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )
tbl_calgary_allsub  <- tbl_regression(
  calgary_any_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )
tbl_edmonton_opioid <- tbl_regression(
  edmonton_opioid_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )
tbl_edmonton_allsub <- tbl_regression(
  edmonton_any_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )

tbl_merge(
  tbls = list(tbl_calgary_opioid, tbl_calgary_allsub, tbl_edmonton_opioid, tbl_edmonton_allsub),
  tab_spanner = c("**Calgary: Opioid**", "**Calgary: Any Substance**",
                   "**Edmonton: Opioid**", "**Edmonton: Any Substance**")
)
  
```

The specificity check compared the estimated post-intervention trend change (`time_after`) between opioid-specific and any-substance death rates for each zone, testing whether the effect observed for Edmonton was specific to opioid, as would be expected if it reflected the SCS's naloxone-based mechanism, or reflected a broader, zone-wide pattern. For Edmonton, the two coefficients were nearly identical (opioid: β = 0.28, p = 0.004; any-substance: β = 0.27, p = 0.004), with overlapping confidence intervals and matching significance. Since naloxone has no plausible mechanism for affecting non-opioid deaths, this near-equivalence across mechanistically distinct outcomes suggests Edmonton's finding more likely reflects a broader confound specific to the zone during this period, rather than an effect attributable to the SCS itself. Calgary showed a similarly consistent pattern across both outcomes (opioid: β = -0.10, p = 0.3; any-substance: β = -0.14, p = 0.14), though neither reached statistical significance in either case.

## Sensitivity Analysis

```{r}
#| label: sensitivity analysis data and model preparation

calgary_data_noSouth  <- calgary_opioid  |> 
  filter(zone != "South")
edmonton_data_noSouth <- edmonton_opioid |> 
  filter(zone != "South")

calgary_noSouth_final <- gls(rate ~ time + zone + intervention + time_after + covid + covid_time,
                             data = calgary_data_noSouth,
                             correlation = corAR1(form = ~ time | zone), method = "REML")
edmonton_noSouth_final <- gls(rate ~ time + zone + intervention + time_after + covid + covid_time,
                              data = edmonton_data_noSouth,
                              correlation = corAR1(form = ~ time | zone), method = "REML")

edmonton_only_opioid <- edmonton_opioid |> 
  filter(zone == "Edmonton")
calgary_only_opioid <- calgary_opioid |> 
  filter(zone == "Calgary")

edmonton_only_opioid_final <- gls(rate ~ time + intervention + time_after + covid + covid_time,
                                  data = edmonton_only_opioid,
                                  correlation = corAR1(form = ~ time), method = "REML")
calgary_only_opioid_final <- gls(rate ~ time + intervention + time_after + covid + covid_time,
                                 data = calgary_only_opioid,
                                 correlation = corAR1(form = ~ time), method = "REML")

```

```{r}
#| label: tbl-4
#| tbl-cap: "Comparison of the estimated trend change (`time_after`) across primary models, South-zone-excluded models, and no-control models"

tbl_noSouth_edmonton <- tbl_regression(
  edmonton_noSouth_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )
tbl_noSouth_calgary <- tbl_regression(
  calgary_noSouth_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )

tbl_edmonton_only <- tbl_regression(
  edmonton_only_opioid_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )

tbl_calgary_only <- tbl_regression(
  calgary_only_opioid_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )

tbl_merge(
  tbls = list(tbl_calgary_opioid, tbl_calgary_only, tbl_noSouth_calgary, tbl_edmonton_opioid, tbl_noSouth_edmonton, tbl_edmonton_only),
 tab_spanner = c("**Calgary Opioid (Controlled model)**", "**Calgary Opioid Excluding South**", "**Calgary Opioid (Single-zone, no controls)**", "**Edmonton Opioid (Controlled model)**", "**Edmonton Opioid Excluding South**", "**Edmonton Opioid (Single-zone, no controls)**")
)  

```


```{r}
#| label: tbl-5
#| tbl-cap: "Comparison of the estimated trend change (`time_after`) between primary models (with comparison zones) and single-zone models"

tbl_edmonton_only <- tbl_regression(
  edmonton_only_opioid_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )
tbl_calgary_only <- tbl_regression(
  calgary_only_opioid_final, 
  include = "time_after", 
  label = list(time_after ~ "Time passed since intervention")
  )

tbl_merge(
  tbls = list(tbl_calgary_opioid, tbl_calgary_only, tbl_noSouth_calgary, tbl_edmonton_opioid, tbl_noSouth_edmonton, tbl_edmonton_only),
 tab_spanner = c("**Calgary Opioid (Controlled model)**", "**Calgary Opioid Excluding South**", "**Calgary Opioid (Single-zone, no controls)**", "**Edmonton Opioid (Controlled model)**", "**Edmonton Opioid Excluding South**", "**Edmonton Opioid (Single-zone, no controls)**")
)  
  
```

Two sensitivity analyses tested whether the primary findings were robust to changes in comparison-group structure. The first excluded South from the comparison zones, given that Lethbridge (South zone) opened its own SCS within weeks of Edmonton's, making it an imperfect control. This is further supported by South zone's parallel pre-trends interaction term showing the largest p-value among the three comparison zones in both models (Calgary: p = 0.076; Edmonton: p = 0.137), though this did not reach conventional statistical significance. The second removed comparison zones entirely, fitting each treated zone's model as a single-group interrupted time series.

Under South-exclusion, Edmonton's trend-change estimate remained significant and slightly strengthened (β = 0.27, p < 0.001), supporting the robustness of the primary finding. Calgary's estimate, however, flipped from non-significant to significant in the protective direction (β = -0.11, p = 0.015). Given the number of specifications tested across this analysis, this result should be interpreted cautiously, as it carries an elevated risk of a false-positive finding arising from multiple comparisons rather than a genuine effect.

Removing comparison zones entirely produced a more substantial change for Edmonton, whose estimate not only lost significance but reversed direction (β = -0.06, p > 0.9). This is not simply a loss of statistical power from a smaller sample; without comparison zones, the model has no external information to distinguish Edmonton's background trend from a genuine SCS-related change, and the two are effectively conflated. This result underscores the importance of the controlled design for Edmonton's primary finding, while also highlighting that the finding's interpretation remains sensitive to the specific comparison structure used. Calgary's estimate under the uncontrolled model remained non-significant (β = -0.28, p > 0.9), consistent with its primary result.

## Summary of Results

Across all analyses, Edmonton's finding proved the more robust of the two. The significant increase in opioid death rate trend following SCS implementation held up under the specificity check (near-identical results for all-substance deaths) and under South-exclusion (effect persisted and slightly strengthened). It disappeared only under the uncontrolled single-zone model, a result attributable to the structural limitation of that specification rather than to reduced statistical power alone. Without comparison zones, the model cannot separate Edmonton's background trend from a genuine SCS-related change. Calgary, by contrast, showed no significant effect in the primary model or the specificity check, and while a significant protective effect emerged under South-exclusion, this result should be treated with caution given the number of specifications tested and the corresponding risk of a false-positive finding. Taken together, these results suggest Edmonton experienced a genuine, if mechanistically unexplained, increase in substance-related mortality following SCS implementation, while Calgary showed no reliable evidence of an effect in either direction.